Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-06-13 SCP2008/paper/metamonads.tex
Have to go through the main generation section

As we argued in \S\ref{CPS}, we will be writing our generic GE
algorithm in the monadic style. For convenience we define the
following type synonyms for our |('p,'v) monad| of Fig.\ref{ourmonad}:
The first is used when we are always generating (abstract) code, while
the second helps when we might generate (abstract) code.
\begin{code}
type ('pc,'v) cmonad = 
  ('p,('a,'v) rep) monad
    constraint
      'p = <state : 's list; answer : ('a,'w) rep>
    constraint
      'pc = <classif : 'a; answer : 'w; state : 's; ..>

type ('pc,'v) omonad = 
  ('p,('a,'v) rep option) monad
    constraint
      'p = <state : 's list; answer : ('a,'w) rep>
    constraint
      'pc = <classif : 'a; answer : 'w; state : 's; ..>
\end{code}

@@

The signature |CONTAINER2D| below specifies that a container must provide
functions |dim1| and |dim2| to extract the dimensions, functions |getL|
to generate container getters, the cloning
generator |copy| and functions that generate code for row and column
swapping. The inclusion of these functions in the signature of all
containers makes it simpler to optimize the relevant functions
depending on the actual representation of the container while not
burdening the users of containers with efficiency details (see
\S\ref{GE-algorithm} for an example of such an optimization).  Note that
this module type is also abstract over the actual representation 
|rep| of the generator data-structure.

\begin{code}
module type CONTAINER2D = sig
  module Dom:DOMAINL
  type contr
  type 'a vc = ('a,contr) rep
  type 'a vo = ('a,Dom.v) rep
  val getL : 'a vc -> ('a,int) rep -> ('a,int) rep -> 'a vo
  val dim1 : 'a vc -> ('a,int) rep
  val dim2 : 'a vc -> ('a,int) rep
  val mapper : ('a vo -> 'a vo) option -> 'a vc -> 'a vc
  val copy : 'a vc -> 'a vc
  val init : ('a,int) rep -> ('a, int) rep -> 'a vc
  val augment : 'a vc -> ('a,int) rep -> ('a, int) rep -> 'a vc ->
                ('a, int) rep -> 'a vc
  val identity : ('a,int) rep -> ('a, int) rep -> 'a vc
  val swap_rows_stmt : 'a vc -> ('a, int) rep -> ('a, int) rep -> 
                       ('a,unit) rep
  val swap_cols_stmt : 'a vc -> ('a, int) rep -> ('a, int) rep -> 
                       ('a,unit) rep
  val row_head : 'a vc -> ('a, int) rep -> ('a, int) rep -> 'a vo
  val col_head_set : 'a vc -> ('a,int) rep -> ('a,int) rep -> 'a vo -> 
            ('a,unit) rep
end
\end{code}

@@

Our sample Gaussian Elimination algorithm, Fig.\ref{fig:GE-algorithm},
demonstrates that tracking determinant is complex: first we define the
variables |det_sign| and |det_magn| used for tracking (lines 22 and
23), then we have to flip the sign when swapping two rows or two
columns or when the matrix found to be singular (lines 32, 37, 59).
The value of the determinant should be updated per each pivoting (line
56). Finally, we convert the tracking state to the resulting
determinant value (lines 64-66). We observe these lines are not
contiguous: the determinant aspect is consulted at several separate
places in the GE algorithm, and the aspect is supposed to keep
state. As other aspects, the determinant aspect is a module of the
following signature:
\begin{code}
module type DETERMINANT = sig
  type tdet = C.Dom.v ref
  type 'a lstate
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) om = ('pc,'v) omonad
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  val decl      : unit -> ('b,unit) lm
  val upd_sign  : unit -> ('b,unit) om
  val zero_sign : unit -> ('b,unit) lm
  val acc_magn  : ('a,C.Dom.v) rep -> (<classif : 'a; ..>,unit) lm
  val get_magn  : unit -> ('b,tdet) lm
  val set_magn  : ('a,C.Dom.v) rep -> (<classif : 'a; ..>,unit) lm
  val fin       : unit -> ('b,C.Dom.v) lm
end
\end{code}
The type |'a lstate| denotes the state being kept as part of the
overall monadic state (tagged as |`TDet|). The signature specifies the
operations of the aspect, such as |decl| for defining the tracking
state, |zero_sign| to note singularity, |set_magn| to set the
magnitude and |fin| to convert the tracking state to the resulting
value, of  the type |C.Dom.v|, the type of the container
elements. Other aspects follow a similar outline.

We have two instances of |DETERMINANT|. The first corresponds to
no determinant tracking, and so all the functions are dummy.
\begin{code}
module NoDet =
  struct
  type tdet = C.Dom.v ref
  type 'a lstate = unit
  let decl ()      = unitL
  let upd_sign ()  = ret None
  let zero_sign () = unitL
  let acc_magn _   = unitL
  let get_magn ()  = ret (liftRef C.Dom.zeroL)
  let set_magn _   = unitL
  let fin () = failwith "Determinant is needed but not computed"
end
\end{code}

The second instance does track the determinant. For integer matrices,
and in general whenever the matrix elements do not form a field, 
the fraction-free update requires tracking some facets of the
determinant, even if we do not output it.
\begin{code}
module AbstractDet =
  struct
  open C.Dom
  type tdet = v ref
  type 'a lstate = ('a,int ref) rep * ('a,tdet) rep
  let decl () = perform
      magn <-- retN (liftRef oneL);    (* track magnitude *)
      sign <-- retN (liftRef Idx.one); (* track the sign: +1, 0, -1 *)
      mo_extend ip (sign,magn);
      unitL
  let upd_sign () = perform
      (sign,_) <-- mo_lookup ip;
      ret (Some (assign sign (Idx.uminus (liftGet sign))))
  let fin = fun () -> perform
      (sign,magn) <-- mo_lookup ip;
      ifM (Logic.equalL (liftGet sign) Idx.zero) (ret zeroL)
      (ifM (Logic.equalL (liftGet sign) Idx.one) (ret (liftGet magn))
          (ret (uminusL (liftGet magn))))
  ...
end
\end{code}
The |decl| method generates two let-bindings for mutable variables
tracking the magnitude and the sign of the determinant, and places the
names of these variables in the monadic state. The |upd_sign| method,
invoked from row- or column-swap generators, retrieves the name of the
sign-accumulating variable and generates the code to update the
sign. The |fin| method retrieves the names of both accumulators and
generates code to compute the final, signed determinant value.  The
generated code closely corresponds to the lines dealing with
|det_magn| and |det_sign| in Fig.~\ref{fig:GE-algorithm}.

@@

More interesting is the aspect of what to return from the LU
algorithm.  One could create an algebraic data type (as was done in
\cite{Carette06}) to encode the various choices: the matrix, the
matrix and the rank, the matrix and the determinant, the matrix, rank
and determinant, and so on. This is wholly unsatisfying as we know
that for any single use, only one of the choices is ever possible, yet
any routine which calls the generated code must deal with these
unreachable options.  Instead we use a module type with an
\emph{abstract} type |res| for the result type; different instances of
the signature set the result type differently. Given below is this
module type and one instantiation, |OutDetRank|, which specifies the
output of a GE algorithm as a 3-tuple |contr * Det.outdet * int| of
the U-factor, the determinant and the rank. That choice of output
corresponds to our sample algorithm in Fig.~\ref{fig:GE-algorithm}.

\begin{code}
module type OUTPUTDEP = sig 
    module PivotRep : PIVOTKIND 
    module Det      : DETERMINANT
end
module type INTERNAL_FEATURES = sig
  module R      : TrackRank.RANK
  module P      : TRACKPIVOT
  module L      : LOWER
end
module type OUTPUT = functor(OD : OUTPUTDEP) -> sig
  module IF : INTERNAL_FEATURES
  type res
  val make_result : 'a wmatrix -> (<classif: 'a;\dots>,res) cmonad
end
module OutDetRank(OD : OUTPUTDEP) = struct
  module IF = struct
    module R   = Rank
    module P   = DiscardPivot
    module L   = NoLower end
  type res = C.contr * C.Dom.v * int
  let make_result m = perform
    det  <-- OD.Det.fin ();
    rank <-- IF.R.fin ();
    ret (Tuple.tup3 m.matrix det rank)
  let _ = OD.Det.fin ()
  let _ = IF.R.fin ()
end
\end{code}

The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
the preflight checks. As we saw in the previous section, 
both instances of |DETERMINANT| contain a |fin ()| function to generate
code representing the computed determinant. Only the |NoDet| does not
track any determinant, and so |fin ()| raises an error. The code 
|let _ = OD.Det.fin ()| in |OutDetRank| invokes this |fin| function, which will
produce the monadic code generating action, or raise an error. We do
not run the action at that time -- we only make sure there is an
action to run. This is another preflight check, to rule out the
semantic error where a user specifies that the determinant should be computed
and returned, and yet specifies the |NoDet| aspect which tracks no
determinant at all.

@@

\subsection{Main generation}
We combine all user-selectable aspects in a record, which serves as a ``keyword
argument'' list to the |GenGE| functor below.
\begin{code}
module type FEATURES = sig
  module Det       : DETERMINANT
  module PivotF    : PIVOT
  module PivotRep  : PIVOTKIND
  module Update    : UPDATE
  module Input     : INPUT
  module Output    : OUTPUT
end
\end{code}

>>>>

As we argued in \S\ref{CPS}, we will be writing our generic GE
algorithm in the monadic style. For convenience we define the
following type synonyms for our |('p,'v) monad| of Fig.\ref{ourmonad}:
The first is used when we are always generating (abstract) code, while
the second helps when we might generate (abstract) code. The helper
type synonym |('pc,'p,'a) cmonad_constraint| is effectively the
abbreviation for a set of constraints imposed on its arguments.
\begin{code}
type ('pc,'p,'a) cmonad_constraint = unit
      constraint
	  'p = <state : 's list; answer : ('a,'w) abstract>
      constraint
          'pc = <classif : 'a; answer : 'w; state : 's; ..>

type ('pc,'v) cmonad = ('p,('a,'v) abstract) monad
      constraint _ = ('pc,'p,'a) cmonad_constraint

type ('pc,'v) omonad = ('p,('a,'v) abstract option) monad
      constraint _ = ('pc,'p,'a) cmonad_constraint 
\end{code}

@@

The signature |CONTAINER2D| below specifies that a container must
provide functions |dim1| and |dim2| to extract the dimensions,
functions |getL| to generate container getters, the cloning generator
|copy| and functions that generate code for row and column
swapping. The inclusion of these functions in the signature of all
containers makes it simpler to optimize the relevant functions
depending on the actual representation of the container while not
burdening the users of containers with efficiency details (see
\S\ref{GE-algorithm} for an example of such an optimization).  We
could have made the type |CONTAINER2D| a functor, parameterized over
|DOMAINL|. The current design simplifies the signature of the generic
GE algorithm at a cost of an extra instantiation step (The
instantiation of the generic algorithm is described in
\S\ref{sec:main-generation}. The expense is worthwhile: currently
OCaml is quite slow when processing functors with complex signatures).
\begin{code}
module type CONTAINER2D = sig
  module Dom:DOMAINL
  type contr
  type 'a vc = ('a,contr) rep
  type 'a vo = ('a,Dom.v) rep
  val getL : 'a vc -> ('a,int) rep -> ('a,int) rep -> 'a vo
  val dim1 : 'a vc -> ('a,int) rep
  val dim2 : 'a vc -> ('a,int) rep
  val mapper : ('a vo -> 'a vo) option -> 'a vc -> 'a vc
  val copy : 'a vc -> 'a vc
  val init : ('a,int) rep -> ('a, int) rep -> 'a vc
  val augment : 'a vc -> ('a,int) rep -> ('a, int) rep -> 'a vc ->
                ('a, int) rep -> 'a vc
  val identity : ('a,int) rep -> ('a, int) rep -> 'a vc
  val swap_rows_stmt : 'a vc -> ('a, int) rep -> ('a, int) rep -> 
                       ('a,unit) rep
  val swap_cols_stmt : 'a vc -> ('a, int) rep -> ('a, int) rep -> 
                       ('a,unit) rep
  val row_head : 'a vc -> ('a, int) rep -> ('a, int) rep -> 'a vo
  val col_head_set : 'a vc -> ('a,int) rep -> ('a,int) rep -> 'a vo -> 
            ('a,unit) rep
end
\end{code}

@@

Our sample Gaussian Elimination algorithm, Fig.\ref{fig:GE-algorithm},
demonstrates that tracking determinant is complex: first we define the
variables |det_sign| and |det_magn| used for tracking (lines 22 and
23), then we have to flip the sign when swapping two rows or two
columns or when the matrix found to be singular (lines 32, 37, 59).
The value of the determinant should be updated per each pivoting (line
56). Finally, we convert the tracking state to the resulting
determinant value (lines 64-66). We observe these lines are not
contiguous: the determinant aspect is consulted at several separate
places in the GE algorithm, and the aspect is supposed to keep
state. As other aspects, the determinant aspect is a module of the
following signature:
\begin{code}
module type DETERMINANT = sig
  type tdet = C.Dom.v ref
  type 'a lstate
  type 'pc pc_constraint = unit
    constraint 'pc = <state : [> `TDet of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint _  = 'pc pc_constraint
  type ('pc,'v) om = ('pc,'v) omonad
    constraint _  = 'pc pc_constraint
  type 'pc nm = ('p,unit) monad
    constraint _ = ('pc,'p,_) cmonad_constraint
    constraint _ = 'pc pc_constraint
  val decl : unit -> 'b nm (* no code is generated *)
  val upd_sign  : unit -> ('b,unit) om
  val zero_sign : unit -> ('b,unit) lm
  val acc_magn  : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val get_magn  : unit -> ('b,tdet) lm
  val set_magn  : ('a,C.Dom.v) abstract -> (<classif : 'a; ..>,unit) lm
  val fin       : unit -> ('b,C.Dom.v) lm
end
\end{code}
The type |'a lstate| denotes the state being kept as part of the
overall monadic state (tagged as |`TDet|). The signature specifies the
operations of the aspect, such as |decl| for initializing the tracking
state, |zero_sign| to note matrix singularity, |set_magn| to set the
magnitude and |fin| to convert the tracking state to the resulting
value, of  the type |C.Dom.v|, the type of the container
elements. Other aspects follow a similar outline.
\oleg{The following is perhaps too particular to mention in the
  paper. AT least we can keep it in comments:
  We stress the difference in return types of \texttt{decl},
  \texttt{upd\_sign} and \texttt{zero\_sign} functions. The former says
  that \texttt{decl} is an action which is executed only for its
  side-effect. It yields no interpretation object (e.g., produces no
  code value). The function \texttt{zero\_sign} always produces a code
  value -- an expression such as assignment that has the type unit. 
  The function
  \texttt{upd\_sign} is the generator that may produce code, or may
  not. The option type lets us avoid generating code such as 
  \texttt{a := b; ()} with useless dummies.}

We have two instances of |DETERMINANT|. The first corresponds to
no determinant tracking, and so all functions are dummy.
\begin{code}
module NoDet = struct
  type tdet = C.Dom.v ref
  type 'a lstate = unit
  let decl ()      = ret ()
  let upd_sign ()  = ret None
  let zero_sign () = unitL
  let acc_magn _   = unitL
  let get_magn ()  = ret (liftRef C.Dom.zeroL)
  let set_magn _   = unitL
  let fin () = failwith "Determinant is needed but not computed"
end
\end{code}

The second instance does track the determinant. For integer matrices,
and in general whenever the matrix elements do not form a field, 
the fraction-free update requires tracking some facets of the
determinant, even if we do not output it.
\begin{code}
module AbstractDet = struct
  open C.Dom
  type tdet = v ref
  type 'a lstate = ('a,int ref) rep * ('a,tdet) rep
  let decl () = perform
      magn <-- retN (liftRef oneL);    (* track magnitude *)
      sign <-- retN (liftRef Idx.one); (* track the sign: +1, 0, -1 *)
      mo_extend ip (sign,magn)
  let upd_sign () = perform
      (sign,_) <-- mo_lookup ip;
      ret (Some (assign sign (Idx.uminus (liftGet sign))))
  let fin = fun () -> perform
      (sign,magn) <-- mo_lookup ip;
      ifM (Logic.equalL (liftGet sign) Idx.zero) (ret zeroL)
      (ifM (Logic.equalL (liftGet sign) Idx.one) (ret (liftGet magn))
          (ret (uminusL (liftGet magn))))
  ...
end
\end{code}
The |decl| method generates two let-bindings for mutable variables
tracking the magnitude and the sign of the determinant, and places the
names of these variables in the monadic state. The |upd_sign| method,
invoked from row- or column-swap generators, retrieves the name of the
sign-accumulating variable and generates the code to update the
sign. The |fin| method retrieves the names of both accumulators and
generates code to compute the final, signed determinant value.  The
generated code closely corresponds to the lines dealing with
|det_magn| and |det_sign| in Fig.~\ref{fig:GE-algorithm}.

\subsection{Other aspects}

For completeness, we briefly describe and show the signatures of the
other aspects. The |RANK| aspect tracks rank; the aspect is a simpler
version of the determinant tracking aspect.  |PIVOTKIND| describes the
representation of the permutation matrix to track permutations
executed in pivoting; |TRACKPIVOT| specifies if these permutations are
tracked at all. |LOWER| tracks the |L| factor when GE algorithm is
used for the in-place LU-decomposition. |PIVOT| is a higher-order
functor abstracting over the pivoting algorithm (full pivoting, row
pivoting, no pivoting, etc). If pivot is found, the aspect should swap
rows and columns appropriately. The swapping may need to be tracked if
we wish to return the permutation matrix. If the determinant is
tracked, the swapping must update the sign of the determinant. The
wapiing also affects the L factor being accumulated in-place. The
|UPDATE| aspect abstracts over the algorithm for updating matrix
elements during row-reductions: full division update or fraction-free
update. Finally, the |INPUT| aspect is an interface aspect abstracting
over the choice of augmented matrix.

\begin{code2}
module type RANK = sig
  type 'a tag_lstate = 'a tag_lstate_
  val decl   : unit -> ('b, int ref) lm
  val succ   : unit -> ('b, unit) lm
  val fin    : unit -> ('b, int) lm
end

module type PIVOTKIND = sig
  type perm_rep
  type 'a ira = ('a, int) rep
  type 'a fra
  type 'a pra = ('a, perm_rep) rep
  val add    : 'a fra -> 'a pra -> 'a pra
  val empty  : 'a ira -> 'a pra
  val rowrep : 'a ira -> 'a ira -> 'a fra
  val colrep : 'a ira -> 'a ira -> 'a fra
end

module type TRACKPIVOT = sig
  type perm_rep
  type 'a ira = ('a, int) rep
  type 'a fra   type 'a pra   type 'a lstate
  type 'pc pc_constraint = unit
    constraint 'pc = <state : [> `TPivot of 'a lstate ]; classif : 'a; ..>
  type ('pc,'v) lm = ('pc,'v) cmonad constraint _  = 'pc pc_constraint
  type ('pc,'a) nm = ('p,unit) monad
    constraint _ = ('pc,'p,'a) cmonad_constraint
    constraint _ = 'pc pc_constraint
  val rowrep : 'a ira -> 'a ira -> 'a fra
  val colrep : 'a ira -> 'a ira -> 'a fra
  val decl   : ('a, int) rep -> ('pc,'a) nm
  val add : 'a fra -> (<classif : 'a; state : [> `TPivot of 'a lstate ]; ..>, unit) omonad
  val fin    : unit -> ('b,perm_rep) lm
end

module type LOWER = sig
  type 'a lstate = ('a, C.contr) rep
  type ('pc,'v) lm = ('pc,'v) cmonad
    constraint 'pc = <state : [> `TLower of 'a lstate ]; classif : 'a; ..>
  val decl   : ('a, C.contr) rep -> (<classif : 'a; ..>, C.contr) lm
  val updt   : 'a C.vc -> ('a,int) rep -> ('a,int) rep -> 'a C.vo -> 
            'a C.Dom.vc -> (<classif : 'a;..>, unit) lm option
  val fin    : unit -> ('a,  C.contr) lm
  val wants_pack : bool
end

module type PIVOT = functor (D: DETERMINANT) -> functor (P: TRACKPIVOT) -> 
          functor (L: LOWER) -> sig
 val findpivot : 'a wmatrix -> 'a curpos -> 
   (<classif : 'a;  state : [> `TDet of 'a D.lstate | `TPivot of 'a P.lstate ]; ..>, 
    C.Dom.v option) cmonad
end

type update_kind = FractionFree | DivisionBased
module type UPDATE = functor(D:DETERMINANT) -> sig
  type 'a in_val = 'a C.Dom.vc
  val update :  'a in_val -> 'a in_val -> 'a in_val -> 'a in_val -> 
      ('a in_val -> ('a, unit) rep) -> ('a, C.Dom.v ref) rep -> 
      (<classif : 'a; ..>, unit) cmonad
  val update_det : 'a in_val -> (<classif : 'a; ..>,unit) D.lm
  val upd_kind : update_kind
end

module type INPUT = sig
  type inp
  val get_input : ('a, inp) rep ->
         (<classif : 'a; ..>, ('a, C.contr) rep * ('a, int) rep * bool) monad
end 
\end{code2}

@@

More interesting is the aspect of what to return from the GE
algorithm.  One could create an algebraic data type (as was done in
\cite{Carette06}) to encode the various choices: the matrix, the
matrix and the rank, the matrix and the determinant, the matrix, rank
and determinant, and so on. This is wholly unsatisfying as we know
that for any single use, only one of the choices is ever possible, yet
any routine which calls the generated code must deal with these
unreachable options.  Instead we use a module type with an
\emph{abstract} type |res| for the result type; different instances of
the signature set the result type differently. Given below is this
module type and one instantiation, |OutDetRank|, which specifies the
output of a GE algorithm as a 3-tuple |contr * Det.outdet * int| of
the U-factor, the determinant and the rank. That choice of output
corresponds to our sample algorithm in Fig.~\ref{fig:GE-algorithm}.

\begin{code}
module type OUTPUTDEP = sig 
    module PivotRep : PIVOTKIND 
    module Det      : DETERMINANT
end
module type INTERNAL_FEATURES = sig
  module R      : TrackRank.RANK
  module P      : TRACKPIVOT
  module L      : LOWER
end
module type OUTPUT = functor(OD : OUTPUTDEP) -> sig
  module IF : INTERNAL_FEATURES
  type res
  val make_result : 'a wmatrix -> (<classif: 'a;...>,res) cmonad
end
module OutDetRank(OD : OUTPUTDEP) = struct
  module IF = struct
    module R   = Rank
    module P   = DiscardPivot
    module L   = NoLower end
  type res = C.contr * C.Dom.v * int
  let make_result m = perform
    det  <-- OD.Det.fin ();
    rank <-- IF.R.fin ();
    ret (Tuple.tup3 m.matrix det rank)
  let _ = OD.Det.fin ()
  let _ = IF.R.fin ()
end
\end{code}

The initialization expressions |OD.Det.fin ()| and |IF.R.fin ()| are
the preflight checks. As we saw in the previous section, both
instances of |DETERMINANT| contain a |fin ()| function to generate
code representing the computed determinant. The instance |NoDet|
however does no tracking, and so |fin ()| raises an error. The code
|let _ = OD.Det.fin ()| in |OutDetRank| invokes this |fin| function,
which will produce the monadic code generating action, or raise an
error. We do not run the action at that time -- we only make sure
there is an action to run. This is another preflight check, to rule
out the semantic error where a user specifies that the determinant
should be computed and returned, and yet specifies the |NoDet| aspect
which tracks no determinant at all.

@@

\subsection{Main generation}\label{sec:main-generation}
We combine all user-selectable aspects in a record, which serves as a ``keyword
argument'' list to the |GenGE| functor below.
\begin{code}
module type FEATURES = sig
  module Det       : DETERMINANT
  module PivotF    : PIVOT
  module PivotRep  : PIVOTKIND
  module Update    : UPDATE
  module Input     : INPUT
  module Output    : OUTPUT
end
\end{code}
